When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents

### 2025-12-05 article.tex
Revise manuscript: add event study analysis, strengthen limitations, update terminology
- Add event study and pre-low-investment analyses to test contrast vs betrayal aversion
- Acknowledge confound between forgiveness and low investment exposure in Limitations
- Change "defection" to "low investment" terminology throughout
- Reframe RS findings: intact detection but impaired recovery in high RS
- Condense Player Ratings section; revise abstract and discussion

%

@@

\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}

@@

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures

@@

Can exposure to forgiving behavior improve cooperative outcomes, and
does Rejection Sensitivity (RS) moderate this effect? This randomized
controlled experiment explores these questions using a novel approach
with Hidden Markov Model (HMM)-based artificial agents in repeated trust
games (RTGs). Participants (N = 206), pre-screened for high or low RS,
interacted with either consistently behaving or more forgiving
co-players, which were in reality HMM agents. Unexpectedly, exposure to
forgiving agents led to reduced cooperation in subsequent interactions.
Control group participants maintained consistent behavior. RS levels
influenced perceptions of the agent's cooperativeness but did not
significantly affect behavioral outcomes, revealing a
perception-behavior dissociation. These findings challenge assumptions
about fostering cooperation through simple exposure to positive
interactions and highlight the importance of accounting for negative
contrast effects when designing such interventions. They also
demonstrate the potential of HMM-based artificial agents for studying
interactive social dynamics, offering a methodological advancement for
future research. Our results suggest the need for nuanced interventions
that gradually shape expectations and behaviors to foster cooperative
outcomes, particularly addressing the discrepancy between perceptions
and behavior in individuals with high rejection sensitivity.

@@

This study reveals that exposing individuals to forgiving behavior in
economic games can unexpectedly decrease their subsequent cooperation.
Interestingly, while people with high rejection sensitivity more
accurately perceived their partners' cooperativeness, this didn't
translate into more cooperative behavior. These findings challenge
assumptions about fostering cooperation and highlight the need for more
nuanced interventions to promote reciprocity, especially for individuals
prone to expecting rejection.

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory (Bowlby, 1978) suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken (Mikulincer, 1998).
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection (Downey et al., 1997; Downey \& Feldman, 1996). RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm (Gao et al., 2017). Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
(Berenson et al., 2009). A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
(Hepp \& Niedtfeld, 2022). The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). Miano et al. (2013) and Richetin et al. (2018) found
that RS mediated the relationship between BPD features and lower trust
appraisal. Abramov et al. (2022) found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between \emph{reciprocity} and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay. Could an ingrained sensitivity to
rejection contribute to reduced pro-social behavior such as a failure to
show reciprocity as a result of perceived rejection by the interaction
partner?

Given that RS may be a manifestation of maladaptive attachment styles,
it's crucial to explore whether exposure to consistently forgiving
interaction partners could help reshape these interpersonal expectations
and behaviors. If RS indeed influences reciprocity in repeated
interactions, could such exposure serve as a corrective experience,
potentially improving interpersonal outcomes? This approach aligns with
attachment theory, which suggests that new, positive relational
experiences can modify internal working models of relationships
(Bowlby, 1988). Research in the fields of behavioral economics
and psychology has also explored how positive social interactions
influence cooperation and pro-social behavior. The use of the repeated
trust game (RTG), a well-established experimental approach, has allowed
for the analysis of the development of trust through ongoing
interactions (Joyce et al., 1995). In this paradigm, cycles of mutual
trust, where each party's trust is reciprocated with trustworthiness,
have the effect of enhancing cooperative behaviors even among
individuals who are initially inclined to be distrustful
(King-Casas et al., 2005). Fowler \& Christakis (2010) studied behavior in
social networks interacting in a public goods game and found that
cooperative behavior tends to cluster, suggesting that exposure to
cooperative peers can lead to more cooperative behavior. Similarly,
research on social learning theory (Bandura, 1977) has long
demonstrated that individuals learn and model the behavior of those
around them, indicating that if someone is consistently exposed to
cooperative and positive individuals, they're likely to emulate this
behavior. These insights highlight that engaging with compassionate and
forgiving others can be an effective method for mitigating deeply
ingrained mistrust and potentially fostering trustworthy behavior.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre Manipulation, Trust-Game Post
Manipulation) design, with repeated measures on the Phase factor (Figure
\ref{fig:HMMPanels}.A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 (Almaatouq et al., 2021). The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

In order to instigate a potential breakdown of trust, thereby allowing
us to probe efforts to repair it, the computerised agent was programmed
to provide a low investment on round 12 (pre-manipulation) and round 13
(post-manipulation). On all other rounds, the investor's actions were
determined by randomly drawing an investment from the state-conditional
distribution, with the state over rounds determined by randomly drawing
the next state from the state-transition distribution as determined from
the net return on the previous round (disregarding the net return
immediately after the pre-programmed low investment rounds).

@@

At the start of the experiment, participants provided informed consent
and were instructed the study would consist of three phases in which
they would face a different other player. Participants were told their
goal was to maximise the number of points in all phases. They were not
told the number of rounds of each phase. Participants were randomly
assigned to either a Control or Manipulation condition. The timeline of
the experiment is shown in Figure \ref{fig:HMMPanels}.A. Phase one
(``pre'') consisted of a 15 round RTG in which participants took the role
of trustee, facing the same investor over all 15 rounds. On each round,
after being informed about the amount sent by the investor participants
decided how much of the tripled investment to return to the investor,
before continuing to the next round. Phase 2 (``exposure'') consisted of
three 7-round RTGs. Participants in the Manipulation condition faced the
forgiving HMM investor and rated the agent on the same attributes as in
the pre-manipulation phase. Those in the Control condition faced the
same human-like HMM agent as in the ``pre'' phase and rated each co-player
on the same attributes. To keep the experience similar to the ``pre''
phase, the agent in the Control condition was also designed to send a
very low investment in round 5 of each of the three games. In the
post-manipulation phase (``post''), participants in both conditions faced
the same human-like HMM as in ``pre'' phase.

@@

{\centering \includegraphics[width=1\linewidth,height=1\textheight]{article_files/figure-latex/HMMPanels-1}

@@

\caption{\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}}\label{fig:HMMPanels}
\end{figure}
\end{landscape}

@@

To test whether participants behaved differently in the RTG after the
manipulation compared to the Control group, we model the percentage
return (percentage of tripled investment returned to investor) using a
linear mixed-effects model with Phase (RTG game pre vs.
post-manipulation), Condition (manipulation vs.~Control), Investment,
and RS (High vs Low RS group) as well as their interactions as fixed
effects, and player-wise random intercepts and slopes for Phase. The
full specification of the statistical model can be found in the
supplement. To test whether participants behaved differently between
conditions in the Exposure phase, we fit a linear mixed effects model to
participants returns only in the exposure phase, with Condition
(manipulation vs.~Control), Investment, and RS (High vs Low RS group) as
well as their interactions as fixed effects, and player-wise random
intercepts. Finally, to test whether the HMM agent's behavior differed
between Phases, Conditions and RS groups, we estimate a linear
mixed-effects model of investments sent by the computerised HMM agent
with Condition, Phase and RS and their interaction as fixed effects, and
a similar random effects structure to the returns model.

The model was estimated using the \texttt{afex} package (Singmann et al., 2022)
in R. More complex models with additional random effects could not be
estimated reliably, and as such the estimated model can be considered to
include the optimal random effects structure
(Matuschek et al., 2017). A similar process was used to establish the
random effects structures of linear mixed-effects models used to analyse
the HMM agent investments as well as the participants' ratings of the
co-players. There is no agreed upon way to calculate effect sizes for
mixed effects models. Instead, we will report on testing differences in
marginal means. For the \(F\)-tests, we used the Kenward-Roger
approximation to the degrees of freedom, as implemented in the R package
``afex''. We Z-transform the Investment variable (subtract the overall
investment mean and divide by overall standard deviation) as centering
is beneficial to interpreting the main effects more easily in the
presence of interactions. To probe significant interactions, we
conducted planned contrasts using the \texttt{emmeans} package in R. Given that
we were testing multiple pre-planned comparisons, we applied the ``Sidak''
correction to control for familywise error rate while maintaining
reasonable statistical power. This approach allowed us to investigate
specific hypotheses about the differential effects of our manipulation
across phases and RS groups, while protecting against inflated Type I
error rates.

@@

Mixed-effects analysis on the percentage returns shows a significant
main effect of Phase (Pre vs.~Post RTG game),
\(F(1, 201.63) = 5.81\), \(p = .017\), with
higher percentage returns in the Pre RTG compared to the Post.
Importantly, we also find an interaction between Condition and Phase
(RTG pre- vs.~post-manipulation),
\(F(1, 201.63) = 4.38\), \(p = .038\).
As shown in Figure \ref{fig:boxPlots}, post-hoc tests confirm that,
contrary to our expectations, there was a decrease in the percentage
returned only in the Manipulation condition, pre - post,
\(\Delta M = 0.03\), 95\% CI \([0.01, 0.05]\), \(t(201.50) = 3.15\), \(p = .002\),
but no change in the Control condition. We find no interaction between
Phase, Condition and RS, suggesting that there was no difference between
RS groups for this interaction.

There was also a significant main effect of Investment,
\(F(1, 5955.67) = 325.35\), \(p < .001\),
such that higher investments were associated with higher percentage
returns indicating positive reciprocity. An Investment by Condition
interaction,
\(F(1, 5955.67) = 13.92\), \(p < .001\),
reflected that returns were more affected by investments in the Control
condition. We also find a three way interaction between Phase,
Investment and RS,
\(F(1, 5864.62) = 7.84\), \(p = .005\),
showing that the differentiated effect of the investment on the
proportion returned by RS group is itself moderated by the Phase (pre-
vs post manipulation). Finally, we find a four-way interaction between
Condition, Phase, Investment and RS
\(F(1, 5864.62) = 9.24\), \(p = .002\).

To examine this four-way interaction, we conducted a contrast analysis
of how the effect of investment on returns changed from pre- to
post-manipulation for different RS groups in both conditions. Starting
with the Manipulation condition, for participants with low RS, the
effect of investment on returns increased significantly from pre- to
post-manipulation,
\(\Delta M = 0.03\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([0.01, 0.05]\), \(t(5881.28) = 3.15\), \(p_\mathrm{\scriptstyle Sidak(3)} = .005\).
This suggests that after the manipulation, low RS participants became
more responsive to their co-player's investments, returning
proportionally more as investments increased. In contrast, for
participants with high RS, the effect of investment on returns decreased
significantly from pre- to post-manipulation,
\(\Delta M = -0.02\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([-0.04, 0.00]\), \(t(5891.81) = -2.67\), \(p_\mathrm{\scriptstyle Sidak(3)} = .023\).
This indicates that high RS participants became less responsive to their
co-player's investments after the manipulation, with smaller increases
in returns as investments increased. The difference in these pre-post
changes between high and low RS groups was significant,
\(\Delta M = -0.05\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([-0.08, -0.02]\), \(t(5887.33) = -4.11\), \(p_\mathrm{\scriptstyle Sidak(3)} < .001\).
This result suggests that the manipulation had significantly different
effects on how low and high RS participants responded to their
co-player's investments.

@@

\caption{Marginal means and distributions of either investments or percentage returns across participants by Phase and Condition. The top panel shows that participants in the Intervention condition returned lower proportions of the multiplied investment received in the second game compared to the first game over all ronds, whilst those in the Control condition sent back similar returns. The bottom panel shows no difference on aggregate of how the HMM invested across Phases and Conditions. RS did not affect returns and investments by Phase and Condition.}\label{fig:boxPlots}
\end{figure}

\subsubsection{Post Defection Trials}\label{post-defection-trials}

Did participants learn to be more forgiving and cooperative after
witnessing the pre-programmed defection by the HMM investor? To explore
this question, we restrict the analysis to the trials following the
pre-programmed defection by the HMM agent in both the ``pre'' (trials 12
to 15) and the ``post'' phases (trials 13 to 15). We fit the same mixed
effects model as for all the trials with the exception of the RS
variable. This is because RS did not show main effects in the previous
model, and also due to the necessity of running a simpler model to
accommodate the low number of trials. We find a significant main effect
of Phase \(F(1, 227.47) = 4.76\), \(p = .030\)
with returns lower in the second game post defection trials compared to
the first. We also find a main effect of Investment
\(F(1, 1296.14) = 156.41\), \(p < .001\)
where participants continued to return higher proportions when receiving
higher investments. Finally, we still find an Investment by Condition
interaction
\(F(1, 5955.67) = 13.92\), \(p < .001\)
showing a lower effects of investment on the Manipulation condition
compared to the Control condition in post-defection trials. However, the
absence of a Condition by Phase effect indicates there was no difference
between conditions on participants' reaction to a one-off low investment
by the co-player.

@@

Figure \ref{fig:plotRatings} shows participants' ratings of each player
they faced by condition and RS group. We will focus on two contrasts to
analyse the ratings by Condition and RS group. The first is between the
rating in the first phase (``pre'') when participants phase the human-like
HMM, and the average rating during the exposure phase (average of
``expo1'', ``expo2'' and ``expo3'') where they either face the forgiving HMM
(Manipulation condition) or the human-like HMM again (Control
condition). The second contrast is between the ``pre'' and ``post'' phases
of the experiment where in both conditions participants face the same
human-like HMM.

\subsubsection{Comparing Pre and Exposure Ratings}\label{comparing-pre-and-exposure-ratings}

>>>>

@@

\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%

@@

@@

Can exposure to forgiving partners improve interpersonal cooperation? Attachment theory suggests positive relational experiences can correct negative internal working models, but individuals high in Rejection Sensitivity (RS)---characterized by anxious expectations of rejection---may struggle to benefit from such experiences. We tested this using a randomized experiment (N = 206) in which participants played repeated trust games with HMM-based artificial agents that simulate human-like trust dynamics. After a baseline game, participants were exposed to either forgiving agents (Manipulation) or standard agents (Control), then played a final game with a standard agent. Contrary to predictions, forgiveness exposure \emph{reduced} subsequent cooperation---participants appeared to perceive the standard post-exposure agent as less cooperative by comparison (a negative contrast effect). Critically, RS moderated specific behavioral patterns but not overall cooperation levels: high RS participants failed to recover cooperation after trust violations and became \emph{less} responsive to partner behavior following exposure, whereas low RS participants showed normal recovery and became \emph{more} responsive. These findings suggest that positive relational experiences do not universally promote cooperation, and that high RS individuals may require interventions targeting their capacity to update expectations rather than simply providing positive experiences.

@@

This study found that exposing people to forgiving partners in economic games unexpectedly decreased their subsequent cooperation---likely because normal partners seemed less cooperative by comparison. Individuals high in rejection sensitivity showed a distinct pattern: while they detected trust violations as readily as others, they failed to restore cooperation afterward and became less responsive to their partner's behavior. In contrast, those low in rejection sensitivity appeared to learn from the positive exposure and became more reciprocal. These findings suggest that simply providing positive social experiences may not benefit everyone equally, and that individuals prone to rejection sensitivity may need targeted support to translate positive experiences into lasting behavioral change.

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory (Bowlby, 1978) suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken (Mikulincer, 1998).
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection (Downey et al., 1997; Downey \& Feldman, 1996). RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm (Gao et al., 2017). Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
(Berenson et al., 2009). A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
(Hepp \& Niedtfeld, 2022). The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). Miano et al. (2013) and Richetin et al. (2018) found
that RS mediated the relationship between BPD features and lower trust
appraisal. Abramov et al. (2022) found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between \emph{reciprocity} and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay.

Given that RS may be a manifestation of maladaptive attachment styles,
it is important to explore whether exposure to consistently forgiving
interaction partners could reshape interpersonal expectations and
behaviors. The \emph{corrective experience hypothesis}, rooted in attachment
theory, suggests that new positive relational experiences can modify
internal working models of relationships (Bowlby, 1988). Research
on social learning (Bandura, 1977) similarly demonstrates that
individuals model the behavior of those around them, and exposure to
cooperative peers promotes cooperative behavior
(Fowler \& Christakis, 2010). In the repeated trust game (RTG) paradigm,
cycles of reciprocated trust enhance cooperative behaviors even among
initially distrustful individuals (King-Casas et al., 2005). This
perspective predicts that exposure to forgiving partners should increase
subsequent cooperation, as participants internalize more positive
expectations about social interactions. Given that high RS individuals
show heightened sensitivity to social cues, they might be particularly
responsive to such corrective experiences.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre-Exposure, Trust-Game Post-Exposure) design, with repeated measures on the Phase factor (Figure
\ref{fig:HMMPanels}.A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 (Almaatouq et al., 2021). The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

In order to instigate a potential breakdown of trust, thereby allowing
us to probe efforts to repair it, the computerised agent was programmed
to provide a low investment on round 12 (pre-exposure) and round 13
(post-exposure). On all other rounds, the investor's actions were
determined by randomly drawing an investment from the state-conditional
distribution, with the state over rounds determined by randomly drawing
the next state from the state-transition distribution as determined from
the net return on the previous round (disregarding the net return
immediately after the pre-programmed low investment rounds).

@@

At the start of the experiment, participants provided informed consent
and were instructed the study would consist of three phases in which
they would face a different other player. Participants were told their
goal was to maximise the number of points in all phases. They were not
told the number of rounds of each phase. Participants were randomly
assigned to either a Control or Manipulation condition. The timeline of
the experiment is shown in Figure \ref{fig:HMMPanels}.A. Phase one
(``pre'') consisted of a 15 round RTG in which participants took the role
of trustee, facing the same investor over all 15 rounds. On each round,
after being informed about the amount sent by the investor participants
decided how much of the tripled investment to return to the investor,
before continuing to the next round. Phase 2 (``exposure'') consisted of
three 7-round RTGs. Participants in the Manipulation condition faced the
forgiving HMM investor and rated the agent on the same attributes as in
the pre-exposure phase. Those in the Control condition faced the
same human-like HMM agent as in the ``pre'' phase and rated each co-player
on the same attributes. To keep the experience similar to the ``pre''
phase, the agent in the Control condition was also designed to send a
very low investment in round 5 of each of the three games. In the
post-exposure phase (``post''), participants in both conditions faced
the same human-like HMM as in ``pre'' phase.

@@

{\centering \includegraphics[width=0.95\linewidth]{article_files/figure-latex/HMMPanels-1}

@@

\caption{A: Experiment timeline. Participants (trustees) played RTGs with HMM investor agents. The investor sends investments (multiplied by 3) and participants decide returns. Conditions differ in exposure phase agents. B-D: The artificial investor is a three-state HMM fitted to human data. B: Investment distributions by latent state. C: Transition probabilities to states at t+1 as a function of net return at t; each panel shows a different starting state. D: Forgiving HMM transitions from low-trust state---unlike the human-like HMM, it always exits low-trust and favors high-trust transitions.}\label{fig:HMMPanels}
\end{figure}
\end{landscape}

@@

We analyzed participants' behavior in the RTG using linear mixed-effects models. First, to examine the effect of the manipulation, we modeled the percentage return (percentage of tripled investment returned to investor) as a function of Phase (RTG game pre vs.~post-exposure), Condition (Manipulation vs.~Control), Investment, and RS (High vs Low RS group), including all interactions as fixed effects. This model included player-wise random intercepts and slopes for Phase. Second, we analyzed behavior during the Exposure phase specifically, modeling returns with Condition, Investment, and RS and their interactions as fixed effects, along with player-wise random intercepts. Third, to verify the consistency of the HMM agent, we modeled the investments sent by the computerized agent using Condition, Phase, and RS and their interactions as fixed effects. To isolate effects occurring prior to any pre-programmed low investment, we also analyzed returns in rounds preceding the low investment trials only (rounds 1-11 in the pre-exposure phase and rounds 1-12 in the post-exposure phase) using the same model specification. Finally, to rigorously assess participants' reactions to and recovery from the specific instance of pre-programmed low investment, we conducted an event study analysis centered on the low investment round (\(t=0\)). We analyzed percentage returns in a three-round window (\(t-1\) to \(t+1\)) using a linear mixed-effects model with Phase, Condition, Time Point, and RS Group as fixed effects. We specifically examined two key behavioral responses: the Drop (change in return from \(t-1\) to the low investment round) and the Recovery (change in return from the low investment round to \(t+1\)). The full specification of all statistical models can be found in the supplement.

All models were estimated using the \texttt{afex} package (Singmann et al., 2022) in R. We determined the random effects structure by starting with the maximal model and simplifying until convergence was achieved, ensuring the optimal structure (Matuschek et al., 2017). A similar process was applied to the models analyzing HMM agent investments and participant ratings. We report differences in marginal means rather than effect sizes, as there is no consensus on effect size calculation for mixed models. \(F\)-tests used the Kenward-Roger approximation for degrees of freedom. The Investment variable was Z-transformed to facilitate the interpretation of main effects in the presence of interactions. Significant interactions were probed using planned contrasts with the \texttt{emmeans} package. We applied the ``Sidak'' correction for multiple comparisons to control the familywise error rate while maintaining statistical power.

@@

Participants returned higher percentages in the Pre phase compared to
the Post phase
(\(F(1, 201.63) = 5.81\), \(p = .017\)). This
effect was moderated by Condition
(\(F(1, 201.63) = 4.38\), \(p = .038\)):
contrary to our expectations, participants in the Manipulation condition
decreased their returns from pre to post
(\(\Delta M = 0.03\), 95\% CI \([0.01, 0.05]\), \(t(201.50) = 3.15\), \(p = .002\)),
while those in the Control condition showed no change (Figure
\ref{fig:boxPlots}). RS did not moderate this Condition × Phase
interaction.

Higher investments elicited higher percentage returns, indicating
positive reciprocity
(\(F(1, 5955.67) = 325.35\), \(p < .001\)).
This relationship was stronger in the Control condition than in the
Manipulation condition
(\(F(1, 5955.67) = 13.92\), \(p < .001\)).
The effect of investment on returns varied by RS group and Phase
(\(F(1, 5864.62) = 7.84\), \(p = .005\)),
and a four-way interaction indicated that these patterns further
differed across Conditions
(\(F(1, 5864.62) = 9.24\), \(p = .002\)).

@@

\caption{Marginal means of percentage returns (top) and HMM investments (bottom) by Phase and Condition. Bars show estimated marginal means; error bars represent 95\% confidence intervals. Participants in the Manipulation condition returned lower proportions post-exposure compared to pre-exposure (** p < .01), while Control participants showed no change. HMM investment did not differ across Phases or Conditions.}\label{fig:boxPlots}
\end{figure}

To examine this four-way interaction, we conducted a contrast analysis
of how the effect of investment on returns changed from pre- to
post-exposure for different RS groups in both conditions (Figure \ref{fig:fourwayInteractionPlot}). Starting
with the Manipulation condition, for participants with low RS, the
effect of investment on returns increased significantly from pre- to
post-phase,
\(\Delta M = 0.03\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([0.01, 0.05]\), \(t(5881.28) = 3.15\), \(p_\mathrm{\scriptstyle Sidak(3)} = .005\).
This suggests that after the manipulation, low RS participants became
more responsive to their co-player's investments, returning
proportionally more as investments increased. In contrast, for
participants with high RS, the effect of investment on returns decreased
significantly from pre- to post-exposure,
\(\Delta M = -0.02\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([-0.04, 0.00]\), \(t(5891.81) = -2.67\), \(p_\mathrm{\scriptstyle Sidak(3)} = .023\).
This indicates that high RS participants became less responsive to their
co-player's investments after the manipulation, with smaller increases
in returns as investments increased. The difference in these pre-post
changes between high and low RS groups was significant,
\(\Delta M = -0.05\), \(95\%\ \mathrm{CI}_\mathrm{\scriptstyle Sidak(3)}\) \([-0.08, -0.02]\), \(t(5887.33) = -4.11\), \(p_\mathrm{\scriptstyle Sidak(3)} < .001\).
This result suggests that the manipulation had significantly different
effects on how low and high RS participants responded to their
co-player's investments.

\begin{figure}[H]
\includegraphics[width=\textwidth]{article_files/figure-latex/fourwayInteractionPlot-1} \caption{Marginal effect of investment on percentage returns by Phase, Condition, and RS group. Bars show estimated slopes (change in returns per SD increase in investment) from the mixed model; error bars represent 95\% confidence intervals. In the Manipulation condition, Low RS participants became more responsive to investments post-exposure (* p < .05), while High RS participants became less responsive (* p < .05). No significant changes were observed in the Control condition.}\label{fig:fourwayInteractionPlot}
\end{figure}

\subsubsection{Returns Prior to Pre-Programmed Low Investment Trials}\label{returns-prior-to-pre-programmed-low-investment-trials}

To distinguish between contrast effects and betrayal aversion as explanations for reduced cooperation in the Manipulation condition, we examined returns in the rounds preceding the pre-programmed low investment. If contrast effects were operating, participants in the Manipulation condition should already show reduced returns before experiencing any low investment in the post-exposure phase. Conversely, if betrayal aversion were the primary mechanism, group differences should only emerge after the low investment.

The Phase \(\times\) Condition interaction was significant in rounds prior to the low investment (\(F(1, 204.05) = 4.86\), \(p = .029\)). Participants in the Manipulation condition significantly decreased their returns from pre- to post-exposure phase even before encountering the low investment (\(\Delta M = 0.03\), 95\% CI \([0.01, 0.06]\), \(t(202.02) = 2.83\), \(p = .005\)), whereas those in the Control condition showed no change (\(\Delta M = 0.00\), 95\% CI \([-0.02, 0.02]\), \(t(201.17) = -0.15\), \(p = .878\)). The four-way interaction also remained significant (\(F(1, 4454.34) = 7.12\), \(p = .008\)), suggesting that the differential responsiveness to investments observed in the full analysis was likewise present before the low investment occurred.

\subsubsection{Reaction to Pre-programmed Low Investment: Event Study Analysis}\label{reaction-to-pre-programmed-low-investment-event-study-analysis}

To understand how participants reacted to and recovered from the pre-programmed low investment, an event study analysis was conducted centered on the low investment round (Figure \ref{fig:eventStudyPlot}). Two behavioral responses were examined: the Drop (change in returns at the low investment round relative to \(t-1\), where negative values indicate reduced returns reflecting punishment of low trust) and Recovery (change in returns at \(t+1\) relative to the low investment round, where positive values indicate restored cooperation).

\begin{figure}[H]
\includegraphics[width=\textwidth]{article_files/figure-latex/eventStudyPlot-1} \caption{Drop and Recovery responses to pre-programmed low investment in the post-exposure phase (pre-exposure phase not shown as no between-condition differences were observed). Drop = change in returns from low investment round minus t-1 (negative values indicate reduced returns); Recovery = change from t+1 minus low investment round (positive values indicate restored returns). Bars show estimated marginal means from the mixed model; error bars represent 95\% confidence intervals. The bracket shows the significant between-condition difference in Recovery for High RS participants. ** p < .01.}\label{fig:eventStudyPlot}
\end{figure}

In the pre-exposure phase, there were no significant differences between conditions in either the Drop (\(M = -0.02\), 95\% CI \([-0.09, 0.05]\), \(t(1773.82) = -0.61\), \(p = .543\)) or Recovery (\(M = 0.00\), 95\% CI \([-0.06, 0.07]\), \(t(1773.56) = 0.11\), \(p = .914\)), confirming that both groups started with equivalent behavioral patterns. In the post-exposure phase, a divergence emerged. The Control group showed a significantly larger Drop than the Manipulation group (\(M = -0.08\), 95\% CI \([-0.15, -0.02]\), \(t(1773.93) = -2.46\), \(p = .014\)), indicating that participants exposed to the forgiving agent showed a blunted immediate reaction to the low investment and did not reduce their returns as sharply. The subsequent Recovery did not differ significantly between conditions overall (\(M = 0.07\), 95\% CI \([0.00, 0.13]\), \(t(1774.02) = 1.92\), \(p = .055\)).

When examining moderation by RS, the pattern appeared to be driven primarily by high RS participants. Low RS participants showed no significant difference in Recovery between conditions (\(M = 0.00\), 95\% CI \([-0.09, 0.10]\), \(t(1774.24) = 0.06\), \(p = .954\)), with both groups displaying modest, similar recovery patterns after the low investment. The larger Drop observed in the Control condition for this group was partly attributable to their elevated cooperation level at \(t-1\).

High RS participants showed a different pattern. In the Control condition, they demonstrated trust repair by significantly increasing their returns after the low investment (\(\Delta M = 0.14\), 95\% CI \([0.08, 0.21]\), \(t(1774.05) = 4.19\), \(p < .001\)). However, those in the Manipulation condition failed to recover, showing no significant increase in returns at \(t+1\) (\(\Delta M = 0.01\), 95\% CI \([-0.05, 0.08]\), \(t(1773.54) = 0.42\), \(p = .676\)). The difference in Recovery between conditions was significant (\(M = 0.13\), 95\% CI \([0.03, 0.22]\), \(t(1773.79) = 2.66\), \(p = .008\)).

In summary, the forgiveness intervention appeared to dampen reciprocal responsiveness, hindering the re-establishment of cooperation following a temporary withdrawal of trust. This effect was more pronounced among high RS individuals. While high RS participants in the Control condition demonstrated active reciprocity by reducing returns sharply when trust was withdrawn and increasing them when trust was restored, those exposed to the forgiving agent exhibited a disengaged pattern characterized by a muted reaction to the low investment and a failure to reinstate high returns afterward.

@@

Figure \ref{fig:plotRatings} shows participants' ratings of co-players across phases. We examined two contrasts: pre-exposure versus exposure phase ratings, and pre-exposure versus post-exposure ratings.
